The fraction of polar aligned circumbinary disks

### 2025-01-14 src/tex/ms.tex
Note that we assume a ring approximates a disk well
use r instead of a

Thus, when the disk lifetime is longer than the alignment timescale, it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}). It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically narrow ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state.

@@

We use the IAS15 Gauss-Radau integrator \citep{reboundias15}, which is a 15th-order variable step size numerical integrator. The maximum time-step this requires is on the order of the binary orbital period, and the integration is therefore much slower than the RKF algorithm, at approximately 2 iterations per second (for $a/a_\text{b}=5$). However, the time required to determine the dynamical state depends on the precession timescale, with some simulation states still undetermined after $10\,000$ outer-mass orbits ($\sim \,$5 seconds). This happens rarely enough that it is most efficient to simply stop the integration, and return an unknown state which must be factored in to our uncertainties. The precession timescale is greatest near the boundary of the precession and libration regions (formally infinite at the boundary, \citealt{farago2010}, see also Figure 16 of \citealt{rabago2024}), and so these boundary simulations are the most likely to be stopped before a dynamical state can be determined. We therefore expect both polar and coplanar simulations to be terminated early at approximately equal rates, and that this method will not introduce bias into our results.

@@

Figure \ref{fig:rkf} shows that, while the RKF and {\sc rebound} simulations produce similar results, they are not identical. Specifically noticeable are points sampled by the MC algorithm that lie outside the RKF libration region (with a higher inclination), but are found by {\sc rebound} to librate. There are also points with the opposite discrepancy -- those that lie in the RKF libration region but produce prograde precession -- but they are fewer and not as noticeable by eye. This discrepancy arises because the secular approximation averages the central potential over the orbital period of the binary, while in this section we are treating the full $n$-body dynamics. The two schemes converge when $a \gg a_\text{b}$, and we can eliminate this discrepancy by increasing $a$ (e.g., $a=25a_\text{b}$ compared to our nominal choice of $5a_\text{b}$). In general, the choice of small $a/a_\text{b}$ changes the shape of the parameter space that allows polar orbits when compared to those found in Section \ref{sec:rk}, but not enough to have a significant effect on $f_\text{polar}$. In fact, this bias is small enough that -- given uncertainties introduced by our limited number of samples -- we cannot distinguish between the polar fractions inferred from the two methods. Because the secular approximation used in Sections \ref{sec:analytic} \& \ref{sec:rk} require significantly lower computational cost, and the differences in results are insignificant, we choose to focus on the secular approximation results in the following sections.

>>>>

Thus, when the disk lifetime is longer than the alignment timescale, it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}). It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically narrow ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state. For simplicity, we assume that the narrow ring model approximates an extended disk sufficiently well. This is justified if warps and breaks are not important to the disk's evolution, though we discuss these effects in the context of our model.

@@

We use the IAS15 Gauss-Radau integrator \citep{reboundias15}, which is a 15th-order variable step size numerical integrator. The maximum time-step this requires is on the order of the binary orbital period, and the integration is therefore much slower than the RKF algorithm, at approximately 2 iterations per second (for $r/a_\text{b}=5$, $r$ is the radius of the ring). However, the time required to determine the dynamical state depends on the precession timescale, with some simulation states still undetermined after $10\,000$ outer-mass orbits ($\sim \,$5 seconds). This happens rarely enough that it is most efficient to simply stop the integration, and return an unknown state which must be factored in to our uncertainties. The precession timescale is greatest near the boundary of the precession and libration regions (formally infinite at the boundary, \citealt{farago2010}, see also Figure 16 of \citealt{rabago2024}), and so these boundary simulations are the most likely to be stopped before a dynamical state can be determined. We therefore expect both polar and coplanar simulations to be terminated early at approximately equal rates, and that this method will not introduce bias into our results.

@@

Figure \ref{fig:rkf} shows that, while the RKF and {\sc rebound} simulations produce similar results, they are not identical. Specifically noticeable are points sampled by the MC algorithm that lie outside the RKF libration region (with a higher inclination), but are found by {\sc rebound} to librate. There are also points with the opposite discrepancy -- those that lie in the RKF libration region but produce prograde precession -- but they are fewer and not as noticeable by eye. This discrepancy arises because the secular approximation averages the central potential over the orbital period of the binary, while in this section we are treating the full $n$-body dynamics. The two schemes converge when $r \gg a_\text{b}$, and we can eliminate this discrepancy by increasing $r$ (e.g., $r=25a_\text{b}$ compared to our nominal choice of $5a_\text{b}$). In general, the choice of small $r/a_\text{b}$ changes the shape of the parameter space that allows polar orbits when compared to those found in Section \ref{sec:rk}, but not enough to have a significant effect on $f_\text{polar}$. In fact, this bias is small enough that -- given uncertainties introduced by our limited number of samples -- we cannot distinguish between the polar fractions inferred from the two methods. Because the secular approximation used in Sections \ref{sec:analytic} \& \ref{sec:rk} require significantly lower computational cost, and the differences in results are insignificant, we choose to focus on the secular approximation results in the following sections.
